---
title: "Multivariable Numerical Algorithms (Gradient & Newton Methods)"
subtitle: "ENGG7518 Research Methods for Engineers: Week 12 & 13"
author: "Prof. Zuduo Zheng"
date: "Semester 2, 2026"
---

::: {.callout-note appearance="simple"}
## Learning Objectives
By the end of this chapter, you should be able to:

1. Formulate the general iterative search procedure in multivariable optimisation, including the roles of Search Direction and Line Search.
2. Execute the **Steepest Descent Method** computationally and explain its zigzagging limitation.
3. Connect Newton’s Method for root-finding (tangent lines) to minimisation (Taylor series quadratic approximation).
4. Formulate and trace **Newton’s Method** using the Hessian matrix, establishing its quadratic convergence speed and descent guarantee.
5. Explain the mechanics of **Quasi-Newton Methods**, applying the Secant Equation and comparing **SR1** and **BFGS** rank-update formulas.
6. Define mathematical conjugacy and explain the memory efficiency of the **Conjugate Gradient (CG) Method**.
:::

## The General Iterative Search Procedure

Analytical optimisation techniques (such as setting the gradient vector to zero, $\nabla f = \mathbf{0}$) often fail in real-world engineering. In practical applications, objective functions are rarely clean analytical polynomials; they frequently depend on complex finite-element simulations, noisy black-box systems, or massive datasets where explicit derivatives are impossible to derive algebraically.

To optimise such systems, we rely on **Numerical Optimisation (Iterative Search Procedures)**.

Starting from an initial guess vector $\mathbf{x}_0 \in \mathbb{R}^n$, numerical algorithms generate a sequence of iterates ($\mathbf{x}_0, \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_k$) that progressively move downhill toward the optimal solution $\mathbf{x}^*$.

At each iteration $k$, the update process relies on a strategy called **Line Search**, which decomposes the multi-dimensional move into two distinct mathematical components:

1. **The Search Direction ($\mathbf{d}_k \in \mathbb{R}^n$):** A directional vector pointing "downhill" in the multi-dimensional decision space.
2. **The Step Length ($\alpha_k \in \mathbb{R}^+$):** A positive scalar determining how far to step along the chosen direction $\mathbf{d}_k$.

The generalised update equation is:
$$ \mathbf{x}_{k+1} = \mathbf{x}_k + \alpha_k \mathbf{d}_k $$

To find the optimal step length $\alpha_k$, the algorithm solves a single-variable (1D) minimisation sub-problem along the chosen ray:
$$ \min_{\alpha_k > 0} \phi(\alpha_k) = f(\mathbf{x}_k + \alpha_k \mathbf{d}_k) $$

```{mermaid}
%%| label: fig-iterative-framework
%%| fig-cap: "General iterative framework for multivariable numerical search algorithms."
flowchart TD
    A["Initial Guess x₀, Tolerance ε"] --> B["Evaluate Gradient ∇f(x_k)"]
    B --> C{"||∇f(x_k)|| ≤ ε?"}
    C -- Yes --> D["Stop: Optimal Solution x*"]
    C -- No --> E["Select Search Direction dₖ"]
    E --> F["Perform Line Search for Step Length αₖ"]
    F --> G["Update Iterate: xₖ₊₁ = xₖ + αₖ dₖ"]
    G --> B
```

### Termination Criteria
Numerical algorithms terminate when one of the following criteria is met:

1. **Convergence Tolerance:** The norm of the gradient falls below a specified tolerance error: $\|\nabla f(\mathbf{x}_k)\| \le \epsilon$, or the relative change in iterates satisfies $\|\mathbf{x}_{k+1} - \mathbf{x}_k\| \le \epsilon$.
2. **Iteration Cap (Backup):** The number of iterations reaches a maximum limit ($k_{\max} = 500$) to prevent infinite loops if the algorithm fails to converge.

Different multivariable algorithms are defined fundamentally by **how they choose the search direction $\mathbf{d}_k$**.

---

## The Steepest Descent Method

The most intuitive choice for a search direction is based on the **Gradient Vector ($\nabla f$)**. 

Because the gradient vector $\nabla f(\mathbf{x}_k)$ points in the direction of steepest *ascent* (the fastest local rate of increase), taking its exact opposite (the negative gradient), yielding the direction of steepest *descent*.

$$\mathbf{d}_k = -\nabla f(\mathbf{x}_k)$$

### The Steepest Descent Algorithm

::: {.callout-note appearance="simple"}
## Algorithm: Steepest Descent (Gradient Descent)
**Step 1: Initialisation**  
Choose an initial point $\mathbf{x}_0 \in \mathbb{R}^n$ and a tolerance error $\epsilon > 0$. Set iteration counter $k = 0$.

**Step 2: Gradient Evaluation & Termination Check**  
Compute the gradient $\nabla f(\mathbf{x}_k)$.  
If $\|\nabla f(\mathbf{x}_k)\| \le \epsilon$, **STOP**. Output $\mathbf{x}^* = \mathbf{x}_k$ as the stationary point.

**Step 3: Search Direction & Line Search**  

1. Set search direction: $\mathbf{d}_k = -\nabla f(\mathbf{x}_k)$.
2. Determine step length $\alpha_k > 0$ by solving the 1D line search:
   $$ \min_{\alpha_k > 0} f(\mathbf{x}_k - \alpha_k \nabla f(\mathbf{x}_k)) $$

3. Update iterate: $\mathbf{x}_{k+1} = \mathbf{x}_k + \alpha_k \mathbf{d}_k$.
4. Increment $k \leftarrow k + 1$ and return to Step 2.
:::

### Worked Example: Steepest Descent with Line Search

::: {.callout-tip}
## Worked Example: 2D Minimisation via Steepest Descent
**Problem:** Minimise $f(x_1, x_2) = 0.5x_1^2 + 2.5x_2^2$ starting from $\mathbf{x}_0 = \begin{bmatrix} 5 \\ 1 \end{bmatrix}$.

**Step 1: Compute Gradient Function**
$$ \nabla f(\mathbf{x}) = \begin{bmatrix} \frac{\partial f}{\partial x_1} \\ \frac{\partial f}{\partial x_2} \end{bmatrix} = \begin{bmatrix} x_1 \\ 5x_2 \end{bmatrix} $$

**Step 2: Evaluate Gradient & Search Direction at $\mathbf{x}_0$**
$$ \nabla f(\mathbf{x}_0) = \begin{bmatrix} 5 \\ 5(1) \end{bmatrix} = \begin{bmatrix} 5 \\ 5 \end{bmatrix} \implies \mathbf{d}_0 = -\nabla f(\mathbf{x}_0) = \begin{bmatrix} -5 \\ -5 \end{bmatrix} $$

**Step 3: Formulate Line Search Sub-Problem**
Express the new iterate as a function of step length $\alpha_0$:
$$ \mathbf{x}_1(\alpha_0) = \mathbf{x}_0 + \alpha_0 \mathbf{d}_0 = \begin{bmatrix} 5 \\ 1 \end{bmatrix} + \alpha_0 \begin{bmatrix} -5 \\ -5 \end{bmatrix} = \begin{bmatrix} 5 - 5\alpha_0 \\ 1 - 5\alpha_0 \end{bmatrix} $$

Substitute into $f(x_1, x_2)$:
$$ \phi(\alpha_0) = 0.5 (5 - 5\alpha_0)^2 + 2.5 (1 - 5\alpha_0)^2 $$

**Step 4: Solve for Step Length $\alpha_0$**
Differentiate $\phi(\alpha_0)$ with respect to $\alpha_0$ and set to zero:
$$ \frac{d\phi}{d\alpha_0} = 0.5 \cdot 2(5 - 5\alpha_0)(-5) + 2.5 \cdot 2(1 - 5\alpha_0)(-5) = 0 $$
$$ -5(5 - 5\alpha_0) - 25(1 - 5\alpha_0) = 0 \implies 150 \alpha_0 = 50 \implies \alpha_0 = \frac{1}{3} $$

**Step 5: Compute Updated Iterate $\mathbf{x}_1$**
$$ \mathbf{x}_1 = \begin{bmatrix} 5 - 5(1/3) \\ 1 - 5(1/3) \end{bmatrix} = \begin{bmatrix} 10/3 \\ -2/3 \end{bmatrix} \approx \begin{bmatrix} 3.333 \\ -0.667 \end{bmatrix} $$
:::

::: {.callout-note}
## Intuitive Metaphor: The Mountain Climber with a Water Bowl
Imagine standing at the base of a fog-covered mountain with a full bowl of water balanced on your head. Your goal is to climb to the summit as fast as possible without spilling water. 

Because of thick fog, you cannot see the peak far away; you only have access to **local information**- the slope under your feet. At each step, you evaluate local slope to choose a direction and step distance. 

Looking back after 500 metres, a step chosen earlier might appear non-optimal in hindsight. However, based on the local information available at that time, it was a completely rational decision!
:::

### Limitations: The "Zigzagging" Problem
While Steepest Descent is reliable and resilient (it will always make progress as long as $\nabla f \ne \mathbf{0}$), it is mathematically **myopic** (shortsighted).

Because it relies *only* on the local slope, it ignores the broader curvature of the surface. When navigating a narrow, elliptical valley, successive steepest descent directions are strictly orthogonal ($\mathbf{d}_k^T \mathbf{d}_{k+1} = 0$). This forces the algorithm to bounce perpendicularly across steep valley walls rather than turning straight down the valley floor. 

Consequently, iterates **zigzag**, leading to a slow **linear convergence rate**.

```{r}
#| label: fig-steepest-descent-zigzag
#| fig-cap: "The 'zigzagging' flaw of the Steepest Descent algorithm in a narrow elliptical valley f(x, y) = 0.5x² + 5y². Consecutive search directions are strictly orthogonal, causing inefficient side-to-side oscillation."
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7.5
#| fig-height: 5

library(ggplot2)

# Grid for contour plot
grid_sd <- expand.grid(x = seq(-0.5, 4.5, length.out = 200),
                       y = seq(-1.2, 1.2, length.out = 200))
grid_sd$z <- 0.5 * grid_sd$x^2 + 5 * grid_sd$y^2

# Compute exact steepest descent trajectory for 6 steps
steepest_descent_path <- function(x0, y0, steps = 6) {
  path <- data.frame(step = 0:steps, x = numeric(steps + 1), y = numeric(steps + 1))
  path$x[1] <- x0
  path$y[1] <- y0
  
  for (k in 1:steps) {
    cx <- path$x[k]
    cy <- path$y[k]
    gx <- cx
    gy <- 10 * cy
    
    num <- gx^2 + 10 * gy^2
    den <- gx^2 + 100 * gy^2
    alpha <- num / den
    
    path$x[k + 1] <- cx - alpha * gx
    path$y[k + 1] <- cy - alpha * gy
  }
  return(path)
}

path_sd <- steepest_descent_path(4, 1, 6)

ggplot() +
  # Contour lines of elliptical valley
  geom_contour(data = grid_sd, aes(x = x, y = y, z = z),
               breaks = c(0.2, 0.8, 2, 4, 7, 11),
               color = "grey60", linewidth = 0.6) +
  # Zigzag path line and points
  geom_path(data = path_sd, aes(x = x, y = y), color = "#D95F02", linewidth = 1.1) +
  geom_point(data = path_sd, aes(x = x, y = y), color = "#D95F02", size = 2.8) +
  # Annotate start and orthogonal steps
  annotate("text", x = 4, y = 1.12, label = expression(paste("Start ", bold(x)[0], " (4, 1)")), color = "#D95F02", size = 3.5) +
  annotate("text", x = 0, y = 0, label = expression(paste("Minimum ", bold(x)*"*", " (0, 0)")), color = "black", size = 3.5, vjust = 1.8) +
  geom_point(aes(x = 0, y = 0), color = "black", size = 3) +
  labs(x = expression(x[1]), y = expression(x[2])) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 11, face = "bold")
  )
```

---

## Newton's Method

To eliminate the zigzagging problem, an algorithm must account for surface curvature using the **Hessian Matrix ($H = \nabla^2 f$)**.

### The Two Perspectives of Newton's Method

Newton's Method can be derived from two complementary mathematical perspectives:

#### 1. The Geometric Root-Finding Perspective
Originally, the Newton-Raphson method solved 1D root-finding problems ($g(x) = 0$) by constructing a tangent line at current guess $x_k$ and finding its $x$-intercept:
$$ x_{k+1} = x_k - \frac{g(x_k)}{g'(x_k)} $$

Because finding a local minimum requires finding a root of the first derivative ($\nabla f(\mathbf{x}) = \mathbf{0}$), we shift all derivatives up by one order:
$$ x_{k+1} = x_k - \frac{f'(x_k)}{f''(x_k)} $$

#### 2. The Minimisation Perspective (Taylor Series Quadratic Model)
In multi-dimensional space, Newton's Method approximates the objective function locally using a **second-order Taylor series expansion** around current iterate $\mathbf{x}_k$:
$$ f(\mathbf{x}_k + \mathbf{p}) \approx m_k(\mathbf{p}) = f(\mathbf{x}_k) + \mathbf{p}^T \nabla f(\mathbf{x}_k) + \frac{1}{2} \mathbf{p}^T \nabla^2 f(\mathbf{x}_k) \mathbf{p} $$

where $\mathbf{p}$ represents the update step vector.

To find the step $\mathbf{p}_k$ that minimises this quadratic model $m_k(\mathbf{p})$, set its gradient with respect to $\mathbf{p}$ equal to zero:
$$ \nabla m_k(\mathbf{p}_k) = \nabla f(\mathbf{x}_k) + \nabla^2 f(\mathbf{x}_k) \mathbf{p}_k = \mathbf{0} $$
$$ \mathbf{p}_k = -[\nabla^2 f(\mathbf{x}_k)]^{-1} \nabla f(\mathbf{x}_k) $$

This yields the **Newton Direction**:
$$ \mathbf{d}_{N, k} = -[H_k]^{-1} \nabla f_k $$

*Automatic Step Length:* Because Newton's Method solves for the exact minimum of the fitted quadratic model, the natural step length $\alpha_k$ is inherently **$1.0$**. Line search is generally unnecessary.

$$ \mathbf{x}_{k+1} = \mathbf{x}_k - [\nabla^2 f(\mathbf{x}_k)]^{-1} \nabla f(\mathbf{x}_k) $$

### The Descent Guarantee
For the Newton step $\mathbf{p}_k$ to genuinely be a *descent direction* ($f(\mathbf{x}_k + \mathbf{p}_k) < f(\mathbf{x}_k)$), the Hessian matrix $\nabla^2 f_k$ **must be Positive Definite** ($\nabla^2 f_k \succ 0$).

If $\nabla^2 f_k \succ 0$, the inner product of the gradient and Newton direction is strictly negative:
$$ \nabla f_k^T \mathbf{p}_k = -\mathbf{p}_k^T \nabla^2 f_k \mathbf{p}_k \le -\sigma_{\min} \|\mathbf{p}_k\|^2 < 0 $$

This mathematically guarantees that the step moves downhill.

### Worked 1D Example: Newton's Method

::: {.callout-tip}
## Worked Example: 1D Minimisation via Newton's Method
**Problem:** Minimise $f(x) = 2x^6 + 3x^4 - 12x$ starting from $x_0 = 1$ with tolerance $\epsilon = 0.0001$.

**Derivatives:**

- $f'(x) = 12x^5 + 12x^3 - 12$
- $f''(x) = 60x^4 + 36x^2$

**Iteration 1 ($k = 0$):**: Evaluate derivatives at $x_0 = 1$: $f'(1) = 12$, $f''(1) = 96$.
- Update step: $p_0 = -\frac{12}{96} = -0.1250$.
- New iterate: $x_1 = 1 - 0.1250 = \mathbf{0.8750}$.
- Error: $|0.8750 - 1.0000| = 0.1250 > 0.0001 \implies$ Continue.

**Iteration 2 ($k = 1$):**: Evaluate derivatives at $x_1 = 0.8750$: $f'(0.8750) = 2.1939$, $f''(0.8750) = 62.7334$.
- Update step: $p_1 = -\frac{2.1939}{62.7334} = -0.03498$.
- New iterate: $x_2 = 0.8750 - 0.03498 = \mathbf{0.8400}$.
- Error: $|0.8400 - 0.8750| = 0.0350 > 0.0001 \implies$ Continue.

By Iteration 4, the algorithm converges to $x^* = \mathbf{0.8376}$ with error $< 0.0001$.
:::

### Strengths and Limitations
* **Strength (Quadratic Convergence):** Newton's Method achieves a **quadratic convergence rate**. Once iterates enter the neighbourhood of the solution, high-precision estimates are obtained in very few steps.
* **Weakness 1 (Sensitivity to Initial Guess):** If $\mathbf{x}_0$ is far from the minimum, parabolic approximations can be inaccurate, causing the method to diverge or oscillate.
* **Weakness 2 (Positive Definite Requirement):** If the Hessian is non-positive definite (e.g., in concave or saddle regions), the Newton direction may point uphill.
* **Weakness 3 (Computational Cost):** Calculating the exact Hessian and solving the linear system $[H]\mathbf{p} = -\nabla f$ requires $\mathcal{O}(n^3)$ floating-point operations. For large-scale problems with thousands of variables, this is computationally prohibitive.

```{r}
#| label: fig-newton-vs-steepest
#| fig-cap: "Comparison between Steepest Descent (zigzagging path in orange) and Newton's Method (direct one-step convergence in teal) on quadratic surface f(x, y) = 0.5x² + 5y²."
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

library(ggplot2)

grid_comp <- expand.grid(x = seq(-0.5, 4.5, length.out = 200),
                         y = seq(-1.2, 1.2, length.out = 200))
grid_comp$z <- 0.5 * grid_comp$x^2 + 5 * grid_comp$y^2

# Steepest descent path
path_sd <- steepest_descent_path(4, 1, 6)
path_sd$Method <- "Steepest Descent (Zigzagging)"

# Newton path (from (4,1) to (0,0) in 1 step)
path_newton <- data.frame(
  step = 0:1,
  x = c(4, 0),
  y = c(1, 0),
  Method = "Newton's Method (Direct Step)"
)

ggplot() +
  # Contour lines
  geom_contour(data = grid_comp, aes(x = x, y = y, z = z),
               breaks = c(0.2, 0.8, 2, 4, 7, 11),
               color = "grey60", linewidth = 0.6) +
  # Steepest Descent Path
  geom_path(data = path_sd, aes(x = x, y = y, color = Method), linewidth = 1.1) +
  geom_point(data = path_sd, aes(x = x, y = y, color = Method), size = 2.8) +
  # Newton Path
  geom_path(data = path_newton, aes(x = x, y = y, color = Method), linewidth = 1.2, linetype = "dashed") +
  geom_point(data = path_newton, aes(x = x, y = y, color = Method), size = 3.5) +
  # Highlights
  scale_color_manual(values = c("Steepest Descent (Zigzagging)" = "#D95F02", 
                               "Newton's Method (Direct Step)" = "#1B9E77")) +
  annotate("text", x = 4, y = 1.12, label = expression(paste("Start ", bold(x)[0], " (4, 1)")), size = 3.5) +
  annotate("text", x = 0, y = -0.15, label = expression(paste("Minimum ", bold(x)*"*", " (0, 0)")), size = 3.5) +
  labs(x = expression(x[1]), y = expression(x[2])) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 11, face = "bold")
  )
```

---

## Quasi-Newton Methods (SR1 and BFGS)

To bypass the $\mathcal{O}(n^3)$ computational bottleneck of computing and inverting exact Hessians while maintaining super-linear convergence, modern algorithms use **Quasi-Newton Methods**.

### The Secant Equation
Instead of computing the exact Hessian matrix from scratch at every step, Quasi-Newton algorithms build a symmetric positive-definite approximation $B_k \approx \nabla^2 f(\mathbf{x}_k)$ (or inverse approximation $H_k \approx [\nabla^2 f(\mathbf{x}_k)]^{-1}$) by accumulating gradient changes over iterations.

Let:

- $\mathbf{s}_k = \mathbf{x}_{k+1} - \mathbf{x}_k$ (change in position)
- $\mathbf{y}_k = \nabla f_{k+1} - \nabla f_k$ (change in gradient)

We require the updated matrix $B_{k+1}$ to satisfy the **Secant Equation**:
$$ B_{k+1} \mathbf{s}_k = \mathbf{y}_k $$

The Quasi-Newton search direction is then: $\mathbf{d}_k = -B_k^{-1} \nabla f_k$.

### Matrix Rank-Update Formulas

Two prominent update formulas are used to construct $B_{k+1}$ from $B_k$:

#### 1. SR1 (Symmetric Rank-One) Formula
Updates the matrix using a rank-one adjustment:
$$ B_{k+1} = B_k + \frac{(\mathbf{y}_k - B_k \mathbf{s}_k)(\mathbf{y}_k - B_k \mathbf{s}_k)^T}{(\mathbf{y}_k - B_k \mathbf{s}_k)^T \mathbf{s}_k} $$

*Characteristics:* Simple and satisfies the Secant Equation, but does **not** guarantee that $B_{k+1}$ remains positive definite.

#### 2. BFGS (Broyden-Fletcher-Goldfarb-Shanno) Formula
BFGS uses a rank-two update and is considered the **gold standard** for non-linear optimisation:
$$ B_{k+1} = B_k - \frac{B_k \mathbf{s}_k \mathbf{s}_k^T B_k}{\mathbf{s}_k^T B_k \mathbf{s}_k} + \frac{\mathbf{y}_k \mathbf{y}_k^T}{\mathbf{y}_k^T \mathbf{s}_k} $$

::: {.callout-important}
## The BFGS Positive Definiteness Guarantee
As long as the initial matrix $B_0$ is positive definite (commonly set to the Identity matrix $I$), and curvature condition $\mathbf{s}_k^T \mathbf{y}_k > 0$ holds, **BFGS mathematically guarantees that every updated matrix $B_{k+1}$ remains Positive Definite**. This ensures stable descent directions at every step.
:::

---

## The Conjugate Gradient (CG) Method

Even storing an $n \times n$ matrix for large-scale engineering problems (such as training deep neural networks or solving 3D PDE systems with millions of variables) exceeds available computer RAM.

The **Conjugate Gradient (CG)** method avoids storing any matrix altogether. Instead, it computes search direction $\mathbf{p}_k$ by taking the negative gradient and adding a fraction of the *previous* search direction:

$$ \mathbf{p}_k = -\nabla f(\mathbf{x}_k) + \beta_k \mathbf{p}_{k-1} $$

where $\beta_k$ is a scalar parameter (e.g., computed via Fletcher-Reeves or Polak-Ribière formulas).

### Mathematical Conjugacy
The directions $\{\mathbf{p}_0, \mathbf{p}_1, \dots, \mathbf{p}_k\}$ are constructed to be **mutually conjugate** with respect to a symmetric positive-definite matrix $A$:

$$ \mathbf{p}_i^T A \mathbf{p}_j = 0 \quad \text{for all } i \neq j $$

```
Geometric Meaning of Conjugacy:
Conjugate directions are orthogonal with respect to surface curvature, preventing search steps from interfering with progress made along previous directions.
```

Because CG implicitly captures curvature through conjugacy without storing matrices, it eliminates the zigzagging flaw of Steepest Descent and converges in at most $n$ iterations for $n$-dimensional quadratic functions, using virtually zero memory overhead.
