---
title: "Mathematical Foundations of Optimisation"
subtitle: "ENGG7518 Research Methods for Engineers: Week 11"
author: "Prof. Zuduo Zheng"
date: "29 July 2026"
---



::: {.callout-note appearance="simple"}
## Learning Objectives
By the end of this chapter, you should be able to:

1. Understand vector operations, inner products, and matrix transformations.
2. Calculate gradient vectors ($\nabla f$) and Hessian matrices ($\mathbf{H}$) for multivariable functions.
3. Test sets and functions for convexity using first- and second-order conditions.
4. Classify matrix definiteness (positive definite, negative definite, indefinite) using Sylvester's Criterion and leading principal minor determinants.
5. Categorise critical points into local minima, local maxima, saddle points, and inflection points.
:::

## Linear Algebra Fundamentals

Before diving into multivariate optimisation, we must establish the basic mathematical vocabulary used to describe multi-dimensional space.

*   **Vector:** A one-dimensional matrix representing the solution of a system of linear equations. In numerical optimisation, unless otherwise stated, a vector refers to a **column vector**. (The transpose of a column vector becomes a row vector).
*   **Matrix:** A rectangular array of numbers. A vector is simply a special type of matrix with only one column or one row. Matrices are essential for storing the higher-order derivatives of complex functions.

## Rates of Change and Curvature: Gradients and Hessians

To find optimal points mathematically, optimisation algorithms must evaluate how functions change in multiple dimensions simultaneously. 

### 1. The Gradient (First Derivative)
For a function of several variables, the standard derivative is replaced by the **Gradient ($\nabla f$)**, which is a vector of its first-order partial derivatives.

**Geometric Interpretation:** 
The gradient always points in the direction of the **greatest rate of increase** of the function. Intuitively, this means that the *negative* gradient ($-\nabla f$) points in the direction where the function value decreases at the fastest rate. 

```{r}
#| label: fig-gradient-contour
#| fig-cap: "Contour plot of $f(x,y) = x^2 + y^2$ showing the gradient vector $\\nabla f(1, 1.5) = [2, 3]^T$ pointing in the direction of steepest ascent, perpendicular to the contour line."
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5

library(ggplot2)

grid <- expand.grid(x = seq(-2, 2, length.out = 100), y = seq(-2, 2, length.out = 100))
grid$z <- grid$x^2 + grid$y^2

# Current location and gradient vector
px <- 1
py <- 1.5
gx <- 2 * px # 2
gy <- 2 * py # 3

arrow_data <- data.frame(
  x = px, y = py,
  xend = px + 0.4 * gx, yend = py + 0.4 * gy
)

ggplot(grid, aes(x = x, y = y)) +
  geom_contour(aes(z = z, color = after_stat(level)), bins = 10, linewidth = 0.8) +
  geom_point(data = data.frame(x = px, y = py), aes(x = x, y = y), color = "#d9534f", size = 3.5) +
  geom_segment(data = arrow_data, aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.3, "cm")), color = "#d9534f", linewidth = 1.2) +
  annotate("text", x = px - 0.35, y = py - 0.15, label = "Current Point (1, 1.5)", fontface = "bold", size = 3.5) +
  annotate("text", x = px + 0.6, y = py + 0.8, label = "Gradient Vector \u2207f", fontface = "bold", color = "#d9534f", size = 3.8) +
  scale_color_viridis_c(name = "f(x,y)") +
  theme_minimal(base_size = 12) +
  labs(
    x = "x",
    y = "y",
    title = "Gradient Vector Points in Direction of Steepest Ascent"
  ) +
  coord_fixed()
```

::: {.callout-tip}
## Example: Evaluating a Gradient Direction
Suppose you are at a current location $(x=2, y=3, z=4)$ in a 3D space, and the general gradient vector of your objective function is given as $\nabla f = (1, 2y, z^2)^T$. 
To find the direction that increases the function's value the fastest from your current location, substitute your coordinates into the gradient:
$$ \nabla f(2,3,4) = (1, 2(3), 4^2)^T = \begin{bmatrix} 1 \\ 6 \\ 16 \end{bmatrix} $$
You must move in the direction of vector $[1, 6, 16]$ to ascend as efficiently as possible. 
:::

*Note on Terminology:* The **Jacobian Matrix ($J$)** is the generalisation of the gradient for vector-valued functions. It is the matrix of all first-order partial derivatives. 

### 2. The Hessian Matrix (Second Derivative)
While the gradient tells us the slope, the **Hessian Matrix ($\mathbf{H}$)** tells us the *curvature*. The Hessian is a square matrix of second-order partial derivatives of a scalar-valued function.

$$ \mathbf{H} = \begin{bmatrix} 
\frac{\partial^2 f}{\partial x_1^2} & \frac{\partial^2 f}{\partial x_1 \partial x_2} & \cdots & \frac{\partial^2 f}{\partial x_1 \partial x_n} \\ 
\frac{\partial^2 f}{\partial x_2 \partial x_1} & \frac{\partial^2 f}{\partial x_2^2} & \cdots & \frac{\partial^2 f}{\partial x_2 \partial x_n} \\
\vdots & \vdots & \ddots & \vdots \\
\frac{\partial^2 f}{\partial x_n \partial x_1} & \frac{\partial^2 f}{\partial x_n \partial x_2} & \cdots & \frac{\partial^2 f}{\partial x_n^2} 
\end{bmatrix} $$

For most real-life continuous functions, the Hessian matrix is **symmetric**. It is a critical component in Newton-type optimisation algorithms, acting as the coefficient of the quadratic term in a local Taylor expansion.

::: {.callout-tip}
## In-Class Exercise: Calculating the Hessian
**Problem:** Obtain the Hessian for the function $f(x,y) = -x^4 + 4(x^2 - y^2) - 3$.

**Step 1: First Derivatives (The Gradient)**

*   $\frac{\partial f}{\partial x} = -4x^3 + 8x$
*   $\frac{\partial f}{\partial y} = -8y$
$$ \nabla f = \begin{bmatrix} -4x^3 + 8x \\ -8y \end{bmatrix} $$

**Step 2: Second Derivatives (The Hessian)**

*   $\frac{\partial^2 f}{\partial x^2} = -12x^2 + 8$
*   $\frac{\partial^2 f}{\partial y^2} = -8$
*   $\frac{\partial^2 f}{\partial x \partial y} = \frac{\partial^2 f}{\partial y \partial x} = 0$
$$ \mathbf{H} = \begin{bmatrix} -12x^2 + 8 & 0 \\ 0 & -8 \end{bmatrix} $$
:::

## Convex Sets and Convexity

### Convex Sets
A set $C$ is **convex** if, for any two points $x_1$ and $x_2$ inside the set, the line segment joining them lies entirely within the set. Mathematically:
$$ \forall x_1 \in C, \forall x_2 \in C \text{ and } \forall \lambda \in [0,1], \quad \lambda x_1 + (1 - \lambda)x_2 \in C $$

```{r}
#| label: fig-convex-sets
#| fig-cap: "Visual comparison of a convex set (left) where any line segment connecting two points stays entirely inside the set, versus a non-convex set (right) where the connecting line segment leaves the set."
#| fig-align: center
#| fig-width: 8
#| fig-height: 4

library(ggplot2)

# Circle (Convex Set)
t <- seq(0, 2*pi, length.out = 100)
convex_df <- data.frame(
  x = cos(t),
  y = sin(t),
  Type = "Convex Set\n(Line stays inside)"
)

# C-Shape / Ring Notch (Non-convex Set)
t_outer <- seq(0.7, 2*pi - 0.7, length.out = 100)
t_inner <- seq(2*pi - 0.7, 0.7, length.out = 100)
nonconvex_df <- data.frame(
  x = c(cos(t_outer), 0.45 * cos(t_inner)),
  y = c(sin(t_outer), 0.45 * sin(t_inner)),
  Type = "Non-Convex Set\n(Line leaves set)"
)

all_sets <- rbind(convex_df, nonconvex_df)

# Points for Convex Set (both inside circle)
pts_convex <- data.frame(
  x = c(-0.5, 0.5), y = c(-0.4, 0.5),
  Type = "Convex Set\n(Line stays inside)"
)

# Points for Non-Convex Set (both inside top/bottom arms of C-shape, segment crosses central hole)
pts_nonconvex <- data.frame(
  x = c(0.7 * cos(1.0), 0.7 * cos(-1.0)),
  y = c(0.7 * sin(1.0), 0.7 * sin(-1.0)),
  Type = "Non-Convex Set\n(Line leaves set)"
)

ggplot(all_sets, aes(x = x, y = y)) +
  geom_polygon(fill = "#2b5c8f22", color = "#2b5c8f", linewidth = 1.2) +
  facet_wrap(~Type) +
  geom_line(data = rbind(pts_convex, pts_nonconvex), aes(x = x, y = y), color = "#d9534f", linetype = "dashed", linewidth = 1) +
  geom_point(data = rbind(pts_convex, pts_nonconvex), aes(x = x, y = y), color = "#d9534f", size = 3) +
  theme_void(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    plot.margin = margin(10, 10, 10, 10)
  ) +
  coord_fixed()
```

### Convex vs. Concave Functions
*   **Linearity:** A property of a function graphically represented as a straight line.
*   **Convexity:** A function is strictly "bowl-shaped". Any line segment drawn between two points on the graph lies *above or on* the graph.
    $$ f(\alpha x + \beta y) \le \alpha f(x) + \beta f(y) \quad \text{where } \alpha + \beta = 1, \; \alpha \ge 0, \; \beta \ge 0 $$

*   **Concavity:** A function is "dome-shaped" or "concave downward". The line segment between two points lies *below or on* the graph.

### The Second Derivative Test for Convexity
To prove a function is convex without graphing it, we look at its second derivative:
> **Rule:** If the Hessian matrix of a function $f$ is **positive semi-definite**, the function $f$ is convex.

::: {.callout-tip}
## In-Class Exercise: Testing Convexity
Determine whether the following single-variable functions are convex:

1.  $f(x) = x^2 \implies f''(x) = 2$. Since $2 > 0$ globally, it is **convex**.
2.  $f(x) = e^x \implies f''(x) = e^x$. Since $e^x > 0$ globally, it is **convex**.
3.  $f(x) = \frac{1}{x} \text{ for } x > 0 \implies f'(x) = -x^{-2} \implies f''(x) = 2x^{-3}$. For $x > 0$, $2/x^3 > 0$, so it is **convex** in this domain.
:::

## Matrix Definiteness and Sylvester’s Criterion

To classify stationary points and test for convexity, we must determine the **definiteness** of the Hessian matrix. For an $n \times n$ symmetric matrix $A$:

1.  **Positive Definite:** $x^T A x > 0$ for all non-zero vectors $x$. ($x=0$ is a unique global minimum).
2.  **Positive Semidefinite:** $x^T A x \ge 0$. ($x=0$ is a global minimum, but not necessarily unique).
3.  **Negative Definite:** $x^T A x < 0$. ($x=0$ is a unique global maximum).
4.  **Negative Semidefinite:** $x^T A x \le 0$. ($x=0$ is a global maximum, but not necessarily unique).
5.  **Indefinite:** The quadratic form is positive in some directions and negative in others. ($x=0$ is a saddle point).

### Testing Definiteness: Leading Principal Minors
To test definiteness without calculating eigenvalues, we evaluate the **Leading Principal Minors**. A leading principal minor is the determinant of the top-left submatrix.

For a $3 \times 3$ matrix, the minors are:

*   $|A_1| = a_{11}$
*   $|A_2| = (a_{11}a_{22}) - (a_{12}a_{21})$
*   $|A_3| = \det(A)$

**The Rules:**

*   $A$ is **Positive Definite** if and only if ALL leading principal minors are strictly positive ($|A_1|>0, |A_2|>0, |A_3|>0$).
*   $A$ is **Negative Definite** if and only if the minors *alternate in sign*, starting with negative ($|A_1|<0, |A_2|>0, |A_3|<0$).
*   If a minor is non-zero but does not fit either sign pattern, the matrix is **Indefinite**.

::: {.callout-tip}
## In-Class Exercise: Matrix Definiteness
Determine the definiteness of the following matrices:

**Matrix 1:** $A = \begin{pmatrix} 2 & 4 \\ 4 & 8 \end{pmatrix}$

*   $|A_1| = 2 > 0$
*   $|A_2| = (2 \times 8) - (4 \times 4) = 16 - 16 = 0$
*   *Result:* Because $|A_1| > 0$ and $|A_2| \ge 0$, the matrix is **Positive Semidefinite**.

**Matrix 2:** $B = \begin{pmatrix} 1 & 2 & 0 \\ 2 & 4 & 5 \\ 0 & 5 & 6 \end{pmatrix}$

*   $|B_1| = 1 > 0$
*   $|B_2| = (1 \times 4) - (2 \times 2) = 0$
*   $|B_3| = 1(24 - 25) - 2(12 - 0) + 0 = -1 - 24 = -25$
*   *Result:* The minors are $1$, $0$, and $-25$. This fits neither the purely positive nor the alternating sign pattern. The matrix is **Indefinite**.
:::

## Categorising Critical Points

With gradients and Hessians understood, we can definitively categorise the critical points of any objective function.

### 1. Stationary Point
For a continuous, differentiable function, a **stationary point** is a point where the gradient is exactly zero ($\nabla f = \mathbf{0}$). The function stops changing at this point. It can be a relative minimum, a relative maximum, or a saddle point.

### 2. Saddle Point (Minimax)
A **saddle point** is a stationary point where the function curves *up* in one direction and *down* in another. It is a relative minimum along one axis (between peaks) and a relative maximum along the crossing axis. 

*   *Geometric view:* Resembles a horse riding saddle or a mountain pass.
*   *Mathematical condition:* Occurs when the Hessian matrix is **Indefinite**.
*   *Optimisation challenge:* Saddle points make locating global minima much harder because algorithms can easily get stuck or slowed down at these coordinates.

::: {.content-visible when-format="html"}
```{r}
#| label: fig-saddle-point-3d
#| fig-cap: "Interactive 3D surface plot of the saddle function $z = x^2 - y^2$. Click and drag to rotate the plot to observe how the surface curves upward along the $x$-axis and downward along the $y$-axis."
#| fig-align: center
#| eval: !expr knitr::is_html_output()

library(plotly)

x <- seq(-2, 2, length.out = 40)
y <- seq(-2, 2, length.out = 40)
z <- outer(x, y, function(x, y) x^2 - y^2)

plot_ly(x = ~x, y = ~y, z = ~z) %>% 
  add_surface(colorscale = "Viridis") %>%
  layout(
    title = list(text = "Interactive 3D Saddle Surface: z = x² - y²"),
    scene = list(
      xaxis = list(title = "x (Curves Up)"),
      yaxis = list(title = "y (Curves Down)"),
      zaxis = list(title = "z = f(x,y)")
    )
  )
```
:::

```{r}
#| label: fig-saddle-point
#| fig-cap: "Contour map of the saddle function $f(x,y) = x^2 - y^2$. The point $(0,0)$ is a relative minimum along the $x$-axis (curves up) and a relative maximum along the $y$-axis (curves down)."
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 5.2

library(ggplot2)

grid_saddle <- expand.grid(x = seq(-2, 2, length.out = 100), y = seq(-2, 2, length.out = 100))
grid_saddle$z <- grid_saddle$x^2 - grid_saddle$y^2

ggplot(grid_saddle, aes(x = x, y = y)) +
  geom_contour_filled(aes(z = z), bins = 12) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "white", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "white", linewidth = 0.8) +
  geom_point(aes(x = 0, y = 0), color = "red", size = 4) +
  annotate("text", x = 0, y = 0.3, label = "Saddle Point (0,0)", color = "white", fontface = "bold", size = 4) +
  annotate("text", x = 1.3, y = 0.15, label = "Curves Up (\u2192 Min)", color = "white", fontface = "italic", size = 3.5) +
  annotate("text", x = 0.3, y = 1.4, label = "Curves Down (\u2191 Max)", color = "white", fontface = "italic", size = 3.5) +
  scale_fill_viridis_d(name = "f(x,y)") +
  theme_minimal(base_size = 12) +
  labs(
    x = "x",
    y = "y",
    title = "Contour Plot of a Saddle Point: f(x,y) = x\u00b2 - y\u00b2"
  ) +
  coord_fixed()
```

### 3. Inflection Point
A concept easily confused with a stationary point is the **inflection point**. An inflection point is where the *curvature* of the curve changes sign (i.e., the second derivative changes sign). 

*   *Key Distinction:* Stationary points are defined by the **first derivative** being zero. Inflection points are defined by the **second derivative**. 
*   An inflection point does *not* necessarily have a zero gradient. However, if an inflection point *is* also a stationary point (gradient is zero), it is technically a saddle point. While saddle points and inflection points are strictly speaking not identical, for many elementary functions, a saddle point acts as an inflection point.